Resolvable Ambiguity

### 2019-07-09 paper/paper.tex
Finished first draft of Orc case study

In Orc, every expression can ``publish'' zero or more values. Orc
defines four combinators for orchestrating these published values:
the parallel (\ocaml{|}), sequential (\ocaml{>}$x$\ocaml{>}),
pruning (\ocaml{<}$x$\ocaml{<}), and ``otherwise'' (\ocaml{;})
combinators. The expression $e_1$ \ocaml{|} $e_2$ runs $e_1$ and
$e_2$ in parallel, publishing any value published by either of
them. The expression $e_1$ \ocaml{>}$x$\ocaml{>} $e_2$ binds each
value published by $e_1$ to $x$ and executes $e_2$, building a
concurrent pipeline. The expression $e_1$ \ocaml{<}$x$\ocaml{<}
$e_2$ binds the \emph{first} value published by $e_2$ to $x$ and
executes $e_1$. Finally, $e_1$\ocaml{;}$e_2$ runs $e_2$ only if
running $e_1$ does not publish any values.

@@

\noindent
Na\"{i}vely composing this definition with another, separately
defined language is likely to introduce ambiguities. For example,
if that language also contains infix operators, the precedence
between these and the Orc combinators will be undefined. With
support for resolvable ambiguity however, we can allow this
ambiguity and let programmers disambiguate expressions using
parentheses. After composing the above definition with a simple
language supporting addition, variables and function calls
(omitted for brevity), parsing the expression
%
``\ocaml{1 + 2 >x> f(x)}''
%
results in the following error message from our tool:

@@

\noindent
In this case, it is likely that the preferred semantics is to have
all operators in the base language bind tighter than the Orc
combinators, and these precedence rules can be added after the
composition without changing the original definitions.

% TODO: Clashes involving ">" or ";"?

>>>>

In Orc, every expression can ``publish'' zero or more values. Orc
defines four combinators for orchestrating these published values:
the parallel (\ocaml{|}), sequential (\ocaml{>}$x$\ocaml{>}),
pruning (\ocaml{<}$x$\ocaml{<}), and ``otherwise'' (\ocaml{;})
combinators. The expression $e_1$ \ocaml{|} $e_2$ runs $e_1$ and
$e_2$ in parallel, publishing any value published by either of
them. The expression $e_1$ \ocaml{>}$x$\ocaml{>} $e_2$ binds each
value published by $e_1$ to $x$ and executes $e_2$ for each such
binding, building a concurrent pipeline. The expression $e_1$
\ocaml{<}$x$\ocaml{<} $e_2$ binds the \emph{first} value published
by $e_2$ to $x$ and executes $e_1$. Finally, $e_1$\ocaml{;}$e_2$
runs $e_2$ only if running $e_1$ does not publish any values.

@@

\noindent
Na\"{i}vely composing this definition with another, separately
defined language is likely to introduce ambiguities. For example,
if that language also contains infix operators, the precedence
between these and the Orc combinators will be undefined. With
support for resolvable ambiguity however, we can allow this
ambiguity and let programmers disambiguate expressions using
parentheses. After composing the above definition with a simple
language supporting addition, variables, and function calls
(language definition omitted for brevity), parsing the expression
%
``\ocaml{1 + 2 >x> f(x)}''
%
results in the following error message from our tool:

@@

\noindent
Because there is no precedence specified between \ocaml{+} and
\ocaml{>x>}, disambiguation is required to specify the order of
operations.
%
In this case, it is likely that the preferred semantics is to have
all operators in the base language bind tighter than the Orc
combinators, and these precedence rules can be added after the
composition without changing the original definitions. Importantly
though, we are not \emph{required} to resolve this ambiguity at
time of composition.

Because of how we defined the syntax of the combinators, and since
parsing with the tool is currently whitespace insensitive, the
multi-character combinators (sequence and pruning) are parsed as
three separate lexical tokens (this is visible in how whitespace
is inserted in the error message above). This means we will also
run into \emph{unresolvable} ambiguities, for example if our base
language includes comparison of numbers with \ocaml{<} and
\ocaml{>}. With such a base language (assuming no associativity
for \ocaml{>}), parsing the expression
%
``\ocaml{42 >x> f(x)}''
%
will result in the following error message:

\begin{lstlisting}[language={[objective]caml}]
Unresolvable ambiguity error with 2 alternatives.
\end{lstlisting}

\begin{minipage}{.3\textwidth}
\begin{lstlisting}[language={[objective]caml}]
Resolvable alternatives:
  ( 42 > x ) > f ( x )
  42 > ( x > f ( x ) )
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}{.5\textwidth}
\begin{lstlisting}[language={[objective]caml}]
Unresolvable alternatives:
  seq
   - int                  gt.orc:1:1-3
   - call                 gt.orc:1:8-12
\end{lstlisting}
\end{minipage}

\noindent
By adding parentheses, a programmer can disambiguate the
expression as two comparisons. However, there is no way for a
programmer to specify that what they want is a sequential
combinator with left and right children being an integer and a
function call respectively. In this case we have at least three
choices to make as language designers:

\begin{itemize}
\item We could decide to forbid the case where two comparisons are
  right next to each other, e.g., \syncon{forbid gt.left = gt} and
  \syncon{forbid gt.right = gt} (where \syncon{gt} is the syncon
  for the greater-than operator).
\item We could change the definition of the parallel combinators
  and define separate lexical tokens for the combinators, e.g.,
  \syncon{token Seq = ">[[:lower:]][[:word:]]*>"}.
\item We could change the syntax of the combinators completely to
  avoid clashes with other operators.
\end{itemize}

The change in the first alternative can be done after composition,
and would allow us to reuse both language definitions without
modifications. In the latter two alternatives, we lose reuse of
the definitions of Orc combinators, but place no additional
restrictions on the base language. The preferred variant is likely
to differ between different compositions and different languages.
